---
title: "Timeseries plot of Swiss residents"
author: "Sara"
date: "`r Sys.Date()`"
output: 
  rmdformats::downcute
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction - what was the task?

As part of the [Data visualisation and communication course](https://www.extensionschool.ch/applied-data-science-communication-visualization), one of the optional tasks after finishing the first of four sub-courses was to find timeseries data and plot it. A very open question! There is so much timeseries data - the vast majority of at least the academic data I have worked with were timeseries. I need to choose something, so somewhat randomly I choose something I haven't worked with before, namely open government data. I want to look at the number of people living in Switzerland - Swiss nationals and foreigners - but additionally show what proportion of the Swiss nationals has/had the right to vote. This could later be compared and contrasted with e.g. the numbers of signatures required for a referendum.

# Getting ready

with package imports
```{r imports, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(scales)
```

## Data sources

all data come from the [Swiss Federal Statistical Office](https://www.bfs.admin.ch/bfs/de/home.html)

- [Share of Swiss nationals 1951 to 2020](https://www.bfs.admin.ch/bfs/de/home/statistiken/bevoelkerung/stand-entwicklung/bevoelkerung.assetdetail.18344337.html)
- [Share of non-Swiss nationals living in Switzerland 1951 to 2020](https://www.bfs.admin.ch/bfs/de/home/statistiken/bevoelkerung/stand-entwicklung/bevoelkerung.assetdetail.18344345.html)
- [share of people allowed to vote](https://www.bfs.admin.ch/bfs/de/home/statistiken/politik.assetdetail.11048403.html)
{Number of signatures required for a referendum described here}(https://hls-dhs-dss.ch/de/articles/010386/2016-07-18/)

## reading and first preparation of data

```{r}
swiss <- read_excel("./data/su-d-01.02.04.02.xlsx", skip=3)
swiss <- swiss %>%
  rename(year = 1, pop_beginning = 2, pop_end = 11) %>% # rename and select columns
  dplyr::select(year, pop_beginning, pop_end) %>%
  filter(str_detect(year, "^[0-9]{4}$")) %>%
  mutate(year = as.integer(year))
glimpse(swiss)
```

```{r}
non_swiss <- read_excel("./data/su-d-01.02.04.03.xlsx", skip = 3)
non_swiss <- non_swiss %>%
  rename(year = 1, pop_beginning = 2, pop_end = 11) %>% # rename and select columns
  dplyr::select(year, pop_beginning, pop_end) %>%
  filter(str_detect(year, "^[0-9]{4}$")) %>%
  mutate(year = as.integer(year))

glimpse(non_swiss)
```

```{r}
voters <- read_excel("./data/su-d-17.02.02.04.01.xlsx", skip = 1)
voters <- voters %>%
  rename(year = 1, population = 3, right_to_vote = 4, voting = 5, proportion_voted = 6) %>%
  select(-2)
glimpse(voters)
```

There are annotations added as e.g. "4)" at the end of an existing data cell, e.g. for year 1939 or the population in 2019. I do not mind losing the 1939 data as some cantons apparently were missing and the numbers take an awkward drop. I'd rather keep the 2019 data though...

```{r}
voters <- voters %>%
  tidyr::separate(population, into = c("population", "addendum"), sep="[[:space:]]")
glimpse(voters)
```

Also, the population number in 2019 is given with ' whereas all the other numbers do not have this!

```{r}
voters <- voters %>%
  select(-addendum) %>%
  filter(str_detect(year, "^[0-9]{4}$")) %>%
  mutate(year = as.integer(year)) %>%
  mutate(population = readr::parse_number(population,
                                          locale = locale(decimal_mark = ".", grouping_mark = "'")))
```

## some sanity check plots

sanity check plots do not need to be nice ;-)

```{r}
ggplot(mapping = aes(x = year)) +
  geom_line(swiss, mapping = aes(y = pop_end)) +
  geom_line(non_swiss, mapping = aes(y = pop_end)) +
  expand_limits(y = 0)
```

```{r}
voters %>%
  ggplot(mapping = aes(x=year)) +
  geom_line(mapping = aes(y = population)) +
  geom_line(mapping = aes(y = right_to_vote)) +
  geom_line(mapping = aes(y = voting))
```

you can easily see when women were added to the pool of people allowed to vote!
plus the plateau of the latest total population value, which apparently was 3 years behind (i.e. really is the value of 2016 rather than 2019)

not yet ideal. more data-munching:

combine swiss and non-swiss into the same tibble and calculate total

```{r}
pop <- swiss %>%
  full_join(non_swiss, by = "year", suffix = c("_ch", "_non_ch")) %>%
  mutate(sum_end = pop_end_ch + pop_end_non_ch) %>%
  mutate(sum_beginning = pop_beginning_ch + pop_beginning_non_ch)
glimpse(pop)
```

but long format would be easier!
```{r}
# let's only take the end-of-year values
pop <- pop %>%
  select(!contains("beginning")) %>%
  rename(ch = pop_end_ch, non_ch = pop_end_non_ch, total = sum_end)

pop_long <- pop %>%
  pivot_longer(cols = !year,names_to = "category", values_to = "population")
```


and quick plot
```{r}
pop_long %>%
  ggplot(mapping = aes(x = year, y = population, color = category)) + 
  geom_line()
```

have this as a stacked area plot
```{r}
pop_long %>%
  filter(category != "total") %>%
  ggplot(mapping = aes(x = year, y = population, fill = category)) + 
  geom_area(color="white")
```

I should convert the category to a factor
```{r}
pop_long <- pop_long %>%
  filter(category != "total") %>%
  mutate(category = factor(category, levels = c("ch", "non_ch")))
```



```{r}
pop_long %>%
  ggplot(mapping = aes(x = year, y = population, fill = fct_rev(category))) + 
  geom_area(size=.5, colour="white", position="stack", alpha=0.8)
```

put the voters in there as well to see how well the total population numbers align:

```{r}
pop_long %>%
  ggplot(mapping = aes(x = year)) + 
  geom_area(mapping = aes(y = population, fill = fct_rev(category)),
            size=.5, colour="white", position="stack", alpha=0.8) +
  geom_line(voters, mapping = aes(y = population), color="black", size = 2) +
  geom_line(voters, mapping = aes(y = voting), size = 2) +
  geom_line(voters, mapping = aes(x = year, y = right_to_vote), size = 2)
```

total population from the voters tibble and from the swiss and non swiss nationals is close enough - take only the latter

filter the voters so that also only data since 1951 is shown
```{r}
voters_to_plot <- voters %>%
  filter(year > 1950)
```

```{r}
pop_long %>%
  ggplot(mapping = aes(x = year)) + 
  geom_area(mapping = aes(y = population, fill = fct_rev(category)),
            size=.5, colour="white", position="stack", alpha=0.8) +
  geom_line(voters_to_plot, mapping = aes(y = right_to_vote), size = 1, color = "black") +
  scale_y_continuous(labels = label_comma(big.mark = "'", scale = 0.001))
```

`{scales}` [package](https://scales.r-lib.org/reference/index.html) is quite for improving the tick labels! apply both a formatting (using the tick as a thousand separator) and a scale (showing 1000 rather than 1 people per unit)

plot with better labels
```{r}
pop_long %>%
  ggplot(mapping = aes(x = year)) + 
  geom_area(mapping = aes(y = population, fill = fct_rev(category)),
            size=.5, colour="white", position="stack", alpha=0.8) +
  geom_line(voters_to_plot, mapping = aes(y = right_to_vote), size = 1, color = "black") +
  scale_y_continuous(labels = label_comma(big.mark = "'", scale = 0.000001)) + 
  labs(title = "Population in Switzerland since 1951",
       x = "",
       y = "population [million people]")
```

join voters to the population tibble so it can be plotted as an area plot as well
but need to split the CH category, otherwise the sum will not be right anymore

```{r}
temp_voters <- voters %>%
  select(year, right_to_vote)

combo_wide <- pop %>%
  filter(year %in% voters$year) %>%
  select(-total) %>%
  left_join(temp_voters, by = "year") %>%
  mutate(ch_no_right = ch - right_to_vote) %>%
  rename(ch_right = right_to_vote) %>%
  select(-ch)

combo_long <- combo_wide %>%
  pivot_longer(cols = c("non_ch", "ch_right", "ch_no_right"),
               names_to = "category",
               values_to = "population") %>%
  mutate(category = fct_relevel(factor(category), c("ch_right", "ch_no_right", "non_ch")))
```


```{r}
combo_long %>%
  ggplot(mapping = aes(x = year, y = population, fill = fct_rev(category))) + 
  geom_area(size=.5, colour="white", position="stack", alpha=0.8) +
  #geom_line(voters_to_plot, mapping = aes(y = right_to_vote), size = 1, color = "black") +
  scale_y_continuous(labels = label_comma(big.mark = "'", scale = 0.000001)) + 
  scale_fill_brewer(labels = c("Not Swiss", "Swiss without right to vote", "Swiss with right to vote")) +
  guides(fill = guide_legend(title = "")) +
  labs(title = "Population in Switzerland since 1951",
       subtitle = "Separated by nationality and the right to vote",
       x = "",
       y = "population [million people]",
       caption = "Source: BfS")
```


```{r}

# get values to be highlighted below
combo_wide <- combo_wide %>%
  mutate(total = non_ch + ch_right + ch_no_right)

# overall population increase
pop_1951 <- combo_wide %>% filter(year == 1951) %>% pull(total)
pop_2019 <- combo_wide %>% filter(year == 2019) %>% pull(total)
total_increase <- (pop_2019 - pop_1951) / pop_1951 * 100

# ch vs non_ch population increase
pop_ch_1951 <- combo_wide %>% filter(year == 1951) %>% mutate(total_ch = ch_right + ch_no_right) %>% pull(total_ch)
pop_non_ch_1951 <- combo_wide %>% filter(year == 1951) %>% pull(non_ch)
pop_ch_2019 <- combo_wide %>% filter(year == 2019) %>% mutate(total_ch = ch_right + ch_no_right) %>% pull(total_ch)
pop_non_ch_2019 <- combo_wide %>% filter(year == 2019) %>% pull(non_ch)
increase_ch <- (pop_ch_2019 - pop_ch_1951) / pop_ch_1951 * 100
increase_non_ch <- (pop_non_ch_2019 - pop_non_ch_1951) / pop_non_ch_1951 * 100

# status 2019
prop_non_swiss_2019 <- combo_wide %>% filter(year==2019) %>% mutate(prop = non_ch/total) %>% pull(prop)

# women get the right to vote
men_only <- combo_wide %>% filter(year == 1967) %>% pull(ch_right)
right_to_vote_1971 <- combo_wide %>% filter(year == 1971) %>% pull(ch_right)

# change in CH since late seventies
pop_ch_1979 <- combo_wide %>% filter(year == 1979) %>% mutate(total_ch = ch_right + ch_no_right) %>% pull(total_ch)
increase_ch_1979_2019 <- (pop_ch_2019 - pop_ch_1979) / pop_ch_1979 *100
```

### Some Summary points

* Between 1951 and 2019, the population living in Switzerland increased from `r format(pop_1951, digits = 0, big.mark = "'", scientific=FALSE)` to `r format(pop_2019, digits = 0, big.mark = "'", scientific=FALSE)`, representing an increase of `r format(total_increase, digits=4)`% - almost doubled!
* The increase is more marked for the foreign population (`r format(increase_ch, digits=4)`%) than for the Swiss population (`r format(increase_non_ch, digits=4)`%).
* This means that in 2019, `r format(prop_non_swiss_2019, digits=4)`% of people living in Switzerland were foreigners - without the right to vote.
* Another very obvious point is the abrupt increase in the number of people with the right to vote in the beginning of the 1970s - this is of course related to women finally (!) being allowed to vote. With this sweeping change, the number of people allowed to vote more than doubled from `r format(men_only, big.mark = "'")` in 1967 to `r format(right_to_vote_1971, big.mark="'")` in 1971.
* Since the women got the right to vote, the proportion of Swiss nationals without the right to vote has decreased. Likely this is related to a change in demographics, with people getting older the younger ones not allowed to vote decrease in proportion.
* In light of the rather large increase in population, it is interesting to note that both the number of signatures required for an optional initiative (to get everyone to vote on a recent decision by parliament), as well as the number of signatures required for a federal populat initiative ('Volksinitiative'), have remained stable since the late seventies. In contrast, the number of Swiss citizens in Switzerland has increased by `r format(increase_ch_1979_2019, digits=4)`%. (To give an exact proportion, the Swiss nationals living abroad would have to be included, but I did not find any numbers older than the last 3 years, so cannot quantify their change over time.)

replot with arrow annotation to women's right to vote
```{r}
combo_long %>%
  ggplot(mapping = aes(x = year, y = population, fill = fct_rev(category))) + 
  geom_area(size=.5, colour="white", position="stack", alpha=0.8) +
  annotate(geom = "curve", x = 1960, xend = 1970, y = 7500000, yend = 3600000,
           curvature = 0.3, arrow = arrow(length = unit(2, "mm")), size=0.7) +
  annotate(geom = "text", x = 1953, y = 8200000, label = "Women got the \nright to vote!", 
           hjust = "left", color="black") +
  scale_y_continuous(labels = label_comma(big.mark = "'", scale = 0.000001)) + 
  scale_fill_brewer(labels = c("Not Swiss", "Swiss without right to vote", "Swiss with right to vote")) +
  guides(fill = guide_legend(title = "")) +
  labs(title = "Population in Switzerland since 1951",
       subtitle = "Separated by nationality and the right to vote",
       x = "",
       y = "population [million people]",
       caption = "Source: Bundesamt für Statistik")
```


Some extra links not used: 
Number of signatures required as comparison

* 100'000 for a federal popular initiative ('Volksinitiative', see [here](https://www.ch.ch/en/political-system/political-rights/initiatives/how-to-launch-a-federal-initiative/)); has been 100'000 since 1977, 50'000 before ([historisches Lexikon der Schweiz: Volksinitiative](https://hls-dhs-dss.ch/de/articles/010386/2016-07-18/))
* 50'000 for an optional referendum (i.e. getting a vote on something parliament has recently decided, see [here](https://www.ch.ch/en/political-system/political-rights/referendums/mandatory-and-optional-referendums/)); has been 50'000 since 1978 ([historisches Lexikon der Schweiz: Referendum](https://hls-dhs-dss.ch/de/articles/010387/2011-12-23/))

to be correct would have to add the Swiss living abroad...[Auslandschweizer in 2019](https://www.pxweb.bfs.admin.ch/sq/1f84b174-583c-45cb-a52e-c6fa8fa8bf78)

Lebenserwartung [hier](https://www.bfs.admin.ch/bfs/de/home/statistiken/bevoelkerung/geburten-todesfaelle/lebenserwartung.assetdetail.18845607.html)
