---
title: "Beautiful maps"
author: "Sara"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: 
  rmdformats::downcute
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(sf)
library(osmdata)
```

# Mapping a city
The task was to choose a city and map at least 5 features with openstreetmap

## Start with Copenhagen! 
with an adjusted bounding box
however, as I found out with Daniel's help, apparently I cannot simply have a vector as a bounding box, but need it to be a matrix...i.e. aiming for this
```{r}
getbb("copenhagen")
```

it's a bit mean, as when I feed it onwards, it looks like a vector:
```{r}
getbb("copenhagen") %>% opq()
```

but when I feed it a vector, the order of the coordinates are wrong!

```{r}
my_bb <- c(55.556724,12.410072,55.716724,12.685072)

opq(my_bb)
```
so have to convert my modified bounding box to a matrix, or switch the order...
```{r}
my_bb <- c(12.410072, 55.556724,12.685072,55.716724)

opq(my_bb)
```
As I will do this again for another city, I'll put the data gathering in a function

```{r}
get_osm_data <- function(bb){
  small_streets <- opq(bb) %>%
  add_osm_feature(
    key = "highway",
    value = c(
      "residential", "living_street",
      "unclassified",
      "service", "footway"
    )
  ) %>%
  osmdata_sf() %>%
  purrr::pluck("osm_lines")

Sys.sleep(1.1)

streets <-  opq(bb) %>%
  add_osm_feature(
    key = "highway",
    value = c(
      "motorway", "primary",
      "secondary", "tertiary"
    )
  ) %>%
  osmdata_sf() %>%
  purrr::pluck("osm_lines")

Sys.sleep(1.1)

railway <- opq(bb) %>%
  add_osm_feature(
    key = "railway",
    value = c(
      "rail"
    )
  ) %>%
  osmdata_sf() %>%
  purrr::pluck("osm_lines")

Sys.sleep(1.1)

metro <- opq(bb) %>%
  add_osm_feature(
    key = "railway",
    value = c(
      "subway"
    )
  ) %>%
  osmdata_sf() %>%
  purrr::pluck("osm_lines")

Sys.sleep(1.1)

water <- opq(bb) %>%
  add_osm_feature(
    key = "natural",
    value = c("bay", "water", "strait")
  )%>%
  osmdata_sf() %>%
  purrr::pluck("osm_multipolygons")

output <- list("small_streets"=small_streets,
            "streets" = streets,
            "railway"= railway,
            "metro"=metro,
            "water"=water)

}
```

```{r}
copenhagen <- get_osm_data(my_bb)
```


```{r}
ggplot() +
  geom_sf(data = copenhagen[["streets"]], color = "#fdbb84", size = 1) +
  geom_sf(data = copenhagen[["small_streets"]], color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = copenhagen[["railway"]],  color = "#f2f2f2", size = 1) +
  geom_sf(data = copenhagen[["metro"]], fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = copenhagen[["water"]], fill = "#3182bd", color = "#3182bd") +
  coord_sf(
    expand = FALSE
  ) +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))
```

hmm, that one water polygon is way bigger than the original bb - need to limit the scales.

```{r}
ggplot() +
  geom_sf(data = copenhagen[["streets"]], color = "#fdbb84", size = 1) +
  geom_sf(data = copenhagen[["small_streets"]], color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = copenhagen[["railway"]],  color = "#f2f2f2", size = 1) +
  geom_sf(data = copenhagen[["metro"]], fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = copenhagen[["water"]], fill = "#3182bd", color = "#3182bd") +
  coord_sf(
    expand = FALSE
  ) +
  scale_x_continuous(limits = c(12.4, 12.69)) +
  scale_y_continuous(limits = c(55.557,55.724)) +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))
```

or crop with `st_crop`
but this again needs yet another format of a bounding box... use a named vector
```{r}
bb <- c("xmin" = 12.4, "xmax" = 12.7, "ymin" = 55.556724, "ymax" = 55.716724)

ggplot() +
  geom_sf(data = st_crop(copenhagen[["streets"]], bb), color = "#fdbb84", size = 1) +
  geom_sf(data = st_crop(copenhagen[["small_streets"]], bb), color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = st_crop(copenhagen[["railway"]], bb),  color = "#f2f2f2", size = 1) +
  geom_sf(data = st_crop(copenhagen[["metro"]], bb), fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = st_crop(copenhagen[["water"]], bb), fill = "#3182bd", color = "#3182bd") +
  #coord_sf(expand = FALSE) +
  #scale_x_continuous(limits = c(12.4, 12.69)) +
  #scale_y_continuous(limits = c(55.557,55.724)) +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))


```

(and no idea why the line of `coord_sf(expand = FALSE)` would make it crash... don't need it anyway.)

```{r}
remove(copenhagen)
```


## Zurich
I like it! I am also going to get stuff from ZH

```{r}
my_bb <- getbb("zurich")
```

```{r}
zh <- get_osm_data(my_bb)
```

```{r}
ggplot() +
  geom_sf(data = zh[["streets"]], color = "#fdbb84", size = 1) +
  geom_sf(data = zh[["small_streets"]], color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = zh[["railway"]],  color = "#f2f2f2", size = 1) +
  geom_sf(data = zh[["metro"]], fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = zh[["water"]], fill = "#3182bd", color = "#3182bd") +
  coord_sf(
    expand = FALSE
  ) +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))
```

same story about the water being larger...

```{r}
ggplot() +
  geom_sf(data = zh[["streets"]], color = "#fdbb84", size = 1) +
  geom_sf(data = zh[["small_streets"]], color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = zh[["railway"]],  color = "#f2f2f2", size = 1) +
  geom_sf(data = zh[["metro"]], fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = zh[["water"]], fill = "#3182bd", color = "#3182bd") +
  coord_sf(
    expand = TRUE
  ) +
  scale_x_continuous(limits = c(8.45, 8.63)) +
  scale_y_continuous(limits = c(47.32, 47.43)) +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))
```

the axis expansion seems to be slightly different in the x and y direction... would like a bit more at the top, so manually tweak

```{r}
ggplot() +
  geom_sf(data = zh[["streets"]], color = "#fdbb84", size = 1) +
  geom_sf(data = zh[["small_streets"]], color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = zh[["railway"]],  color = "#f2f2f2", size = 1) +
  geom_sf(data = zh[["metro"]], fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = zh[["water"]], fill = "#3182bd", color = "#3182bd") +
  coord_sf(
    expand = TRUE
  ) +
  scale_x_continuous(limits = c(8.45, 8.63)) +
  scale_y_continuous(limits = c(47.32, 47.435)) +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))
```

but better with crop:
```{r}
bb <- c("xmin" = 8.45, "xmax" = 8.63, "ymin" = 47.32, "ymax" = 47.43)
ggplot() +
  geom_sf(data = st_crop(zh[["streets"]], bb), color = "#fdbb84", size = 1) +
  geom_sf(data = st_crop(zh[["small_streets"]], bb), color = "#fee8c8", alpha = 0.6) +
  geom_sf(data = st_crop(zh[["railway"]], bb),  color = "#f2f2f2", size = 1) +
  geom_sf(data = st_crop(zh[["metro"]], bb), fill = "#b2b2b2", size = 0.6) +
  geom_sf(data = st_crop(zh[["water"]], bb), fill = "#3182bd", color = "#3182bd") +
  theme_void() +
  theme(plot.background = element_rect(fill = "#282828"))

```

there we go. and now I can go into artistic colors with the [met brewer](https://github.com/BlakeRMills/MetBrewer) package

```{r}
library(MetBrewer)
```

```{r}
plot_zurich <- function(pal, bb){
p <- ggplot() +
  geom_sf(data = st_crop(zh[["streets"]], bb), color = pal[1], size = 1) +
  geom_sf(data = st_crop(zh[["small_streets"]], bb), color = pal[2], alpha = 0.6) +
  geom_sf(data = st_crop(zh[["railway"]], bb),  color = pal[3], size = 1) +
  geom_sf(data = st_crop(zh[["metro"]], bb), fill = pal[4], size = 0.6) +
  geom_sf(data = st_crop(zh[["water"]], bb), fill = pal[5], color = pal[5]) +
  theme_void() +
  theme(plot.background = element_rect(fill = pal[6],
                                       color = pal[6]))

p
}
```

```{r}
pal <- met.brewer(name="Tiepolo", n=6, type = "discrete")
pal
```


```{r}
p <- plot_zurich(pal)
p
```

this is fun! I'll do some more:

```{r}
pal <- met.brewer(name="Lakota", n=6, type = "discrete")
pal
```

```{r}
q <- plot_zurich(pal)
q
```

also a bit dark, but with more variety. Funny to see the lake in green!

```{r}
pal <- met.brewer(name="Juarez", n=6, type = "discrete")
pal
```

```{r}
r <- plot_zurich(pal)
r
```

quite similar...
```{r}
pal <- met.brewer(name="Homer2", n=6, type = "discrete")
pal
```

```{r}
s <- plot_zurich(pal)
s
```

I like this one :-)


I'll save the ones I like best
the plotting already takes a long time...

```{r}
ggsave("zurich1.png", plot = p, dpi=300)
ggsave("zurich2.png", plot = q, dpi=300)
ggsave("zurich3.png", plot = r, dpi=300)
ggsave("zurich4.png", plot = s, dpi=300)
```

